---
title: "Final Business Analysis Churn"
author: "Hunter Davis"
date: "2025-05-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Final Busines Analysis Churn Data

```{r Load, echo=TRUE, results='hide', warning=FALSE}
# Load required libraries
library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(factoextra)
library(FactoMineR)
```

```{r EDA}
# 1. Data Loading and Exploration --------------------------------------------

data <- read.csv("~/AMU/Dats201/Week 8/Telecom_Customer_Churn_Dataset.csv")
str(data)
summary(data)
head(data)
colnames(data)
colSums(is.na(data))

data$CustomerID <- NULL

# Convert categorical variables to factors
data$HasTechSupport <- as.factor(data$HasTechSupport)
data$IsSeniorCitizen <- as.factor(data$IsSeniorCitizen)
data$Churn <- as.factor(data$Churn)
data$NumOfProducts <- as.factor(data$NumOfProducts)

# Data Visualization
ggplot(data, aes(x=Churn, fill=Churn)) + 
  geom_bar() + 
  ggtitle("Churn Distribution")

ggplot(data, aes(x=MonthlyCharges, fill=Churn)) +
  geom_density(alpha=0.5) +
  ggtitle("Monthly Charges by Churn Status")

numeric_vars <- c("Age",
                  "AnnualIncome",
                  "MonthlyCharges",
                  "TotalCharges",
                  "TenureMonths")

data %>%
  pivot_longer(cols = all_of(numeric_vars), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = Churn, y = value, fill = Churn)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y") +
  labs(title = "Numeric Variables by Churn Status",
       x = "Churn",
       fill = "Churn Status")
```

### Key Findings (EDA):

#### EDA

* No missing values (colSums(is.na(data)) = 0 for all columns)
* Balanced Churn rate: 51% Yes, 49% No
* Numeric variable ranges:
  + Age: 18-69 years (mean 43.82)
  + Annual Income: $30232-$79971 (mean $54910)
  + Monthly Charges: $30.01-$119.85 (mean $74.95)
  + Total Charges: $500-$7991.70 (mean $4239)
  + Tenure: 1-71 Months (mean 35.89)
<br>

#### Preprocessing

* Converted categorical variables to factors:
  + HasTechSupport (No/Yes)
  + IsSeniorCitizen (0/1)
  + Churn (No/Yes)
  + NumOfProducts (1-4)

```{r Predict}
# 2. Predictive Modeling --------------------------------------------------

# Split data
set.seed(1)
trainIndex <- createDataPartition(data$Churn, p=0.7, list=FALSE)
train <- data[trainIndex,]
test <- data[-trainIndex,]

# Logistic Regression
logit_model <- glm(Churn ~ ., data=train, family="binomial")
summary(logit_model)

# Decision Tree
tree_model <- rpart(Churn ~ ., data=train, method="class")
summary(tree_model)
rpart.plot(tree_model, main="Decision Tree for Customer Churn")

```

### Key findings (Model Evaluation):

#### Logistic Regression

* Significant predictors (p<0.05)
  + Annual Income (coefficient = -1.366e-05, p = 0.0103)
  + Is Senior Citizen (coefficient = -0.3274, p = 0.0326)
* Annual income: 
  + Estimate = -1.366e-05 (log-odds per $1)
  + For $10,000 increase: -1.366e-05 × 10,000 = -0.1366 (log-odds)
  + Odds Ratio = exp(-0.1366) = 0.872
  + Interpretation: Each $10K income increase leads to a 12.8% reduction in churn odds (1 - 0.872 = 0.128)
* Is Senior Citizen:
  + Estimate = 0.3274 (log-odds)
  + Odds Ratio = exp(0.3274) = 1.387
  + Interpretation: Seniors have 38.7% higher churn odds than non-seniors
* Non-Significant Predictors (p<0.05)
  + Age (p=0.6405): No significant age effect after accounting for other factors
  + Monthly Charges (p=0.5611): Pricing not significant when income is considered
  + Tenure Months (p=0.7373): Surprisingly, tenure length doesn't independently predict churn
  + Number of Products: All categories (2,3,4) vs baseline (1) show no significant differences (p=0.75, 0.74, 0.35)
  + Tech Support (p=0.9094): No protective effect found
<br>
The logistic Model shows moderate performance with an AIC of 980.38 and residual deviance of 958.38 on 690 degrees of freedom.
<br>

#### Decision Tree

* Key Splits
  + Primary Split:
    + Annual Income < $48,664 (Right branch: 253 obs, 57.7% churn)
    + Higher-income customers (>$48K) have lower churn risk (45.1% churn)
  + Secondary Split:
    + Age < 24.5: Younger customers in high-income group show 64% retention
    + Total Charges < $931: Customers with lower cumulative spending churn more
    + Monthly Charges < $104: Pricing sensitivity appears at this threshold
  + Deep Splits:
    + Node 15: Customers with Age >=53.5 & Income <$46.5K have 85.7% churn rate
    + Node 232: Customers with Age <56.5 & Total Charges <$2,944 show 90% retention
<br>
* Variable importance:
  + Total Charges (36%)
  + Annual Income (32%)
  + Age (19%)
  + Monthly Charges (10%)
  + Tenure (2%)
  


```{r Resample}
# Resampling Methods ------------------------------------------------------

# Set up cross-validation
ctrl <- trainControl(method = "cv",
                     number = 10,
                     savePredictions = TRUE,
                     classProbs = TRUE,
                     summaryFunction = twoClassSummary)

# Logistic Regression with CV
logit_cv <- train(Churn ~ ., data=train,
                  method = "glm",
                  family = "binomial",
                  trControl = ctrl,
                  metric = "ROC")
print(logit_cv)

# Decision Tree with CV
tree_cv <- train(Churn ~ ., data=train,
                 method = "rpart",
                 trControl = ctrl,
                 tuneLength = 10,
                 metric = "ROC")
print(tree_cv)
plot(tree_cv, main="Decision Tree Tuning Results")

# Model comparison
models <- list(Logistic = logit_cv, DecisionTree = tree_cv)
resamples <- resamples(models)
summary(resamples)
bwplot(resamples, metric = "ROC", main="Model Comparison by ROC")

```

### Key findings (Model resampling):

#### Cross-validation Results

* Logistic Regression
  + ROC: 0.515
  + Sensitivity: 0.530
  + Specificity: 0.451
* Decision Tree (optimal cp = 0.030)
  + ROC: 0.546
  + Sensitivity: 0.604
  + Specificity: 0.489
  
#### Comparison

The decision tree model shows a better overall performance with a higher ROC, sensitivity, and specificity. The box plot reinforces this by visualizing each models spread and mean value.

```{r Unsupervised}
# Unsupervised Learning ---------------------------------------------------

# PCA Analysis
pca_data <- data %>% select(all_of(numeric_vars)) %>% scale()
pca_result <- PCA(pca_data, graph=FALSE)
summary(pca_result)

# PCA Visualization
fviz_eig(pca_result, addlabels=TRUE, main="Scree Plot - Variance Explained")
fviz_pca_var(pca_result, col.var="contrib",
             gradient.cols=c("#00AFBB", "#E7B800", "#FC4E07"),
             repel=TRUE, title="Variable Contributions to PCs")

# K-means Clustering with PCA components
pca_scores <- as.data.frame(pca_result$ind$coord)
colnames(pca_scores) <- paste0("PC", 1:ncol(pca_scores))

# Determine optimal clusters
fviz_nbclust(pca_scores, kmeans, method="wss")+
  geom_vline(xintercept = 3, linetype = 2) +

# Perform clustering
set.seed(123)
kmeans_result <- kmeans(pca_scores, centers=3, nstart=25)
data$Cluster <- as.factor(kmeans_result$cluster)

# Visualize clusters
fviz_cluster(kmeans_result, data=pca_scores, 
             geom="point",
             ellipse.type="convex",
             palette="jco",
             main="Customer Segments from PCA")
```

```{r Insights}
# Analysis and Insight ----------------------------------------------------

# Cluster analysis
cluster_summary <- data %>%
  group_by(Cluster) %>%
  summarise(
    Count = n(),
    ChurnRate = mean(Churn == "Yes"),
    AvgMonthlyCharge = mean(MonthlyCharges),
    AvgTenure = mean(TenureMonths),
    AvgAge = mean(Age),
    AvgIncome = mean(AnnualIncome)
  )

print(cluster_summary)

# Key churn factors
varImp(logit_cv)
varImp(tree_cv)
```

### Key findings (PCA and K-means):

#### Principal Component Analysis

* The scree plot showed the first two components explained most variance
* Variable contribution plot indicated:
  + PC1 strongly influenced by tenure and charges
  + PC2 influenced by monthly charges, income, and age

#### K-means Clustering
  
* Cluster 1 (30.1% of customers):
  + Highest tenure (47.9 months)
  + Older customers (avg age 50.1)
  + Moderate churn rate (47.5%)
* Cluster 2 (36.4% of customers):
  + Younger customers (avg age 29.1)
  + Moderate tenure (32.6 months)
  + Highest churn rate (51.1%)
* Cluster 3 (33.5% of customers):
  + Oldest customers (avg age 54.2)
  + Shortest tenure (28.7 months)
  + High monthly charges ($79.80)
  + Moderate churn rate (49.9%)
    
## Summary and Recomendations

### Critical Churn Factors

* Financial Factors:
  + Lower annual income customers are more likely to churn
  + Customers with higher monthly charges show higher churn tendencies
* Demographic Factors:
  + Senior citizens have significantly higher churn risk
  + Younger customers (Cluster 2) show the highest churn rates
* Service Factors:
  + Number of products didn't show strong predictive power
  + Tech support availability wasn't a significant factor

### Strategic Recommendations

* Targeted Retention Programs:
  + Implement income-based pricing strategies for lower-income customers
  + Create senior citizen-specific service plans with added benefits
  + Develop loyalty programs for younger customers (Cluster 2)
* Service Optimization:
  + Review pricing structure for high monthly charge customers
  + Analyze why high-tenure customers (Cluster 1) still churn at 47.5% rate
* Proactive Engagement:
  + Implement early warning systems using the decision tree model
  + Focus outreach on customers matching high-risk cluster profiles
* Further Research:
  + Investigate why tech support isn't reducing churn as expected
  + Collect additional data on service quality and customer satisfaction
  + Explore more complex models to improve predictive accuracy

The analysis suggests that a segmented, data-driven approach focusing on financial and demographic factors will be most effective in reducing customer churn for this telecom company.